The impact of gas bulk rotation on the Lyman $\alpha$ line

### 2014-08-11 paper/ms.tex
new ms version

\begin{document}

@@

In this paper we express a photon's frequency in terms of the
dimensionless variable $x\equiv (\nu -\nu_a)/\Delta\nu_\alpha$, where
$\nu_{\alpha}=2.46\times 10^{15}$ Hz is the Ly$\alpha$ resonance
frequency,  $\Delta\nu_{\alpha} \equiv
\nu_{\alpha}\sqrt{2kT/m_pc^2}\equiv \nu_av_{\rm th}/c $ is the Doppler
broadening of the line which depends on the neutral gas temperature
$T$ scattering the radiation or equivalently the thermal velocity
$v_{\rm th}$ of the atoms.  For the temperature $T=10^4$K used in our
radiative transfer calculations the thermal velocity is
$v_{\rm th}=12.8$\kms.

@@

If the viewing angle is aligned with the rotation axis, $|\mu|\sim
1$, the \ly line keeps in most of the cases a double peak with minor
changes as the rotational velocity starts to increase. 
However, for a line of sight perpendicular to the rotation axis,
$|\mu|\sim 0$, the impact of  rotation is larger.  
The double peak readily transforms into a single peak.

@@

This is a remarkable result: while the rotation axis defines preferential direction, the
integrated flux is the same for all viewing angles in the range of parameters explored in this paper. {\bf This can be understood from the fact that
{\it radiative transfer inside a sphere that undergoes solid-body rotation proceeds identical as inside a static sphere}: we can draw a line between any two atoms within the rotating cloud, and their relative velocity along this line is zero (apart from the relative velocity as a result of random thermal motion), irrespective of the rotation velocity of the cloud. This relative velocity is what is relevant for the radiative transfer\footnote{This point can be further illustrated by considering the path of individual photons: let a photon be emitted at line center ($x=0$), in some random direction ${\bf k}$, propagate a distance that corresponds to $\tau_0=1$, scatter fully coherently (i.e. $x=0$ after scattering in the gas frame) by 90$^{\circ}$, and again propagate a distance that corresponds to $\tau_0=1$. The position where the photon scatters next does {\it not} depend on the rotation of the cloud, nor on ${\bf k}$.}.}

@@

\begin{figure*}
\begin{center}
    \includegraphics[width=0.45\textwidth]{f12.pdf}
    \includegraphics[width=0.45\textwidth]{f13.pdf}
\end{center}
\caption{Logarithm of the average number of scatterings as function of
  $\mu$ (left) and the maximum rotational velocity $V_{\rm
max}$ (right). The left panel shows the behaviour for $\tau=10^{5}$ and
  $V_{max}=300$\kms as a function of $\left|\cos\theta\right|$; the independence of $N_{\rm scatt}$ with $\mu$ and $V_{\rm max}$ is present in all models.
\label{fig:Nscatt} }
\end{figure*}

@@

%In order to gain a deeper understanding of these results  we make 2D
%histograms for the number of scatterings as a function of the outgoing
%dimensionless frequency $x$. In Fig. \ref{fig:Nscatt2D} we show these 
%results in the case $\tau_{\rm H}=10^5$ for the
%static case and $V_{\rm max}=300$\kms. The upper (lower) panels show the
%results for the homogeneous (central) source distribution. The color
%scale is logarithmic in the number of photons at a certain value
%$x-N_{\rm scatt}$.

%The top right panel of Fig. \ref{fig:Nscatt2D} (homogeneous sources,
%high rotational velocity) supports our hypothesis about
%the photo-sphere in the homogeneous distribution. In this case most of
%the photons that left with $x\sim 0$ have escaped with less than $10$
%scatterings. This also explains the decrease in the average number of
%scatterings observed in Fig. \ref{fig:Nscatt}. We have also veryfied
%that these photons are indeed emitted closer to the surface than to
%the center of the sphere.

%However, for a central distribution the situation is quite different
%(lower panels in Fig. \ref{fig:Nscatt2D}). In this case the number
%of scatterings remains high, in the order of the optical depth, but
%the two peaks do get closer to each other. The most probable
%physical picture is that each scattering, due to the bulk velocity of
%the gas, is inefficient in driving the photon outside the line
%center.

@@

%An alternative - more quantitative way - to view this is   to consider the path of an individual photon: say the photon travels  a distance corresponding to $\tau=1$, scatters by $90$ degrees, then travels another distance corresponding to $\tau=1$. The point where the photon ends up is independent of the assumed rotation velocity of the cloud.

>>>>

%@arxiver{fig3.pdf,fig11a.pdf, fig11b.pdf} 
\begin{document}

@@

In this paper we express a photon's frequency in terms of the
dimensionless variable $x\equiv (\nu -\nu_a)/\Delta\nu_{\rm D}$, where
$\nu_{\rm \alpha}=2.46\times 10^{15}$ Hz is the Ly$\alpha$ resonance
frequency,  $\Delta\nu_{\rm D} \equiv
\nu_{\alpha}\sqrt{2kT/m_pc^2}\equiv \nu_av_{\rm th}/c $ is the Doppler
broadening of the line which depends on the neutral gas temperature
$T$ or equivalently the thermal velocity
$v_{\rm th}$ of the atoms. We also use the parameter $a$ to define the
relative line width as $a=\Delta\nu_{\alpha}/2\Delta\nu_{\rm D}$,
where $\Delta\nu_{\alpha}$ is the intrinsic linewidth. For the
temperature $T=10^4$K used in our radiative transfer calculations the
thermal velocity is $v_{\rm th}=12.8$\kms.

@@

If the viewing angle is aligned with the rotation axis, $|\mu|\sim
1$, the \ly line keeps a double peak with minor
changes in the morphology as the rotational velocity increases. 
However, for a line of sight perpendicular to the rotation axis,
$|\mu|\sim 0$, the impact of  rotation is larger.  
The double peak readily transforms into a single peak.

@@

This is a remarkable result: while the rotation axis defines preferential direction, the
integrated flux is the same for all viewing angles in the range of parameters explored in this paper. This can be understood from the fact that
{\it radiative transfer inside a sphere that undergoes solid-body
  rotation proceeds identical as inside a static sphere}: we can draw
a line between any two atoms within the rotating cloud, and their
relative velocity along this line is zero (apart from the relative
velocity as a result of random thermal motion), irrespective of the
rotation velocity of the cloud. This relative velocity is what is
relevant for the radiative transfer\footnote{This point can be further
  illustrated by considering the path of individual photons: let a
  photon be emitted at line center ($x=0$), in some random direction
  ${\bf k}$, propagate a distance that corresponds to $\tau_0=1$,
  scatter fully coherently (i.e. $x=0$ after scattering in the gas
  frame) by 90$^{\circ}$, and again propagate a distance that
  corresponds to $\tau_0=1$. The position where the photon scatters
  next does {\it not} depend on the rotation of the cloud, nor on
  ${\bf k}$.}

@@

\begin{figure*}
\begin{center}
    \includegraphics[width=0.45\textwidth]{f12.pdf}
    \includegraphics[width=0.45\textwidth]{f13.pdf}
\end{center}
\caption{Logarithm of the average number of scatterings as function of
  $\mu$ (left) and the maximum rotational velocity $V_{\rm
max}$ (right). The left panel shows the behaviour for $\tau=10^{5}$ and
  $V_{max}=300$\kms as a function of $\left|\cos\theta\right|$. In the
  right panel the continous (dashed) lines represent the results for
  the central (homogeneous) model. The independence of $N_{\rm scatt}$ with $\mu$ and $V_{\rm max}$ is
  present in all models.  
\label{fig:Nscatt} }
\end{figure*}

@@

@@
